Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-07 papers/termination/paper.tex
Experiments

To evaluate our algorithm, we implemented two tools: one to generate a
termination specification from a C program, and one to synthesise a proof
from the specification.  We ran the resulting termination prover on 47
benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of $180\,s$, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are given in Figure~\ref{fig:experiments}.

It should be noted that the comparison here is imperfect, since {\sc ARMC}
is solving a slightly different problem -- it checks whether the program
under analysis would terminate if run with unbounded integer variables,
while we are checking whether the program terminates with fixed-width
integer variables.  This means that {\sc ARMC}'s verdict differs from ours
in 3 cases (due to the differences between integer and bit-vector
semantics).  There are a further 5 cases where our tool is able to find a
proof and {\sc ARMC} cannot, which we believe is due to our more expressive
proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to
find a termination proof.  All of these cases have nested loops, which are
easy for us to handle, but can be difficult for provers that enumerate
lassos.

On the other hand, {\sc ARMC} is \emph{much} faster than our tool.  While
this difference can partly be explained by much more engineering time being
invested in {\sc ARMC}, we feel that the difference is probably inherent to
the difference in the two approaches -- our algorithm is complete and
bit-vector accurate, which comes at the cost of efficiency.

@@

\begin{figure}
\centering
\scriptsize
\begin{tabular}{|l|@{}c@{}||@{}c@{}|r||@{}c@{}|r|r|}
\hline
          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{Us} \\
Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time & \#It.\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}

>>>>

To evaluate our algorithm, we implemented two tools: one to generate a
termination specification from a C program, and one to synthesise a proof
from the specification.  We ran the resulting termination prover,
named {\sc Juggernaut}, on 47
benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of $180\,s$, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are given in Figure~\ref{fig:experiments}.

It should be noted that the comparison here is imperfect, since {\sc ARMC}
is solving a slightly different problem -- it checks whether the program
under analysis would terminate if run with unbounded integer variables,
while we are checking whether the program terminates with fixed-width
integer variables.  This means that {\sc ARMC}'s verdict differs from ours
in 3 cases (due to the differences between integer and bit-vector
semantics).  There are a further 5 cases where our tool is able to find a
proof and {\sc ARMC} cannot, which we believe is due to our more expressive
proof language.  In 3 cases, {\sc ARMC} times out while our tool is able to
find a termination proof.  2 of these cases have nested loops and the
third has an infinite number of terminating lassos.  This is not a problem for us,
but can be difficult for provers that enumerate lassos.

On the other hand, {\sc ARMC} is \emph{much} faster than our tool.  While
this difference can partly be explained by much more engineering time being
invested in {\sc ARMC}, we feel that the difference is probably inherent to
the difference in the two approaches -- our algorithm is complete and
bit-vector accurate, which comes at the cost of efficiency.
{\sc Juggernaut} is slow, but unstoppable in its
termination abilities if you are willing to wait for it.

@@

Of the 47 benchmarks used, 2 used nonlinear operations in the program (loop6 and loop11),
and 5 had nested loops (svcomp6, svcomp12, svcomp18, svcomp40, svcomp41).
{\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
It solves 2 of the 5 nested loops in less than 30s, but times out on the other
3.  Of these 3 timeouts, 2 (svcomp40 and svcomp41) are due program constants that
are too large to be handled with our default settings.  Increasing
our default word width from 3 bits to 5 bits allowed us to prove termination
for both these benchmarks in less than 30s (but these results are not shown in
the table).

In conclusion, while they show that our tool is not yet production ready,
these experiments confirm our conjecture that synthesis can be used
effectively to prove termination and non-termination.  In particular,
for programs with nested loops, nonlinear arithmetic and complex
termination arguments, the versatility of unrestricted synthesis
is very valuable.

\begin{figure}
\centering
\scriptsize
\begin{tabular}{|l|@{}c@{}||@{}c@{}|r||@{}c@{}|r|r|}
\hline
          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{\sc Juggernaut} \\
Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time & \#It.\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}
